---
title: "Introduction to Statistics with R"
subtitle: "Session R01: Basics and Diagrams"
author: "Marvin Schmitt"
output:
  beamer_presentation:
    slide_level: 2
    toc: no
    theme: Berlin
    includes:
      in_header: ../../config-files/notes-config.tex
header-includes: \setbeamertemplate{footline}[]{}
---




```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 3, fig.width = 6)

library(tidyverse)
library(ggplot2)
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
set.seed(42)
```

## Prequel: The pipe operator `%>%`
If you load the `tidyverse` library, you can use the `%>%` pipe. Instead of `function_b(function_a(x))` we can write 

`x %>% function_a() %>% function_b()`
```{r}
x = 10
round(log(x), 3)
x %>% log() %>% round(3)
```

The ` %>% ` operator works like a **pipe** and passes the left-hand-side as the first argument to the function on the right-hand-side.

## Sampling from a distribution
`rnorm(n=n, mean=mu, sd=sigma)` draws `n` samples from the normal distribution $\mathcal{N}(\mu, \sigma)$:
```{r}
IQ_values = rnorm(n=200, mean=87, sd=10) %>% round()
print(IQ_values)
```

## Descriptive statistics
```{r}
mean(IQ_values)          # Mean
var(IQ_values)           # Variance
sd(IQ_values)            # Standard Deviation sigma
min(IQ_values)           # Minimum
max(IQ_values)           # Maximum
max(IQ_values) - min(IQ_values)  # Range
```

## Plots: Histogram in base `R`
```{r}
hist(IQ_values, breaks = 20,
     main="IQ Distribution", xlab = "Intelligenzquotient"
     )
```

## Plots: `ggplot2` introduction

- `ggplot2` is a modern library to generate publication-ready plots. 
- When you are visualizing data, it should usually be your first choice. 
- The `ggplot2` syntax is modular and different from the base `R` syntax. 
-`ggplot2` works best on data frames, so let's turn `x` into a data frame:

```{r}
df_IQ = data.frame(IQ_values)
```

## Plots: `ggplot2` histogram

```{r}
ggplot(data=df_IQ, aes(x=IQ_values)) +
  geom_histogram(fill="pink", color="black") +
  labs(title="IQ distribution", x="IQ")
```

*** 


```{r}
ggplot(data=df_IQ, aes(x=IQ_values)) +
  geom_histogram(binwidth=20, fill="pink", color="black")
```

***

```{r}
ggplot(data=df_IQ, aes(x=IQ_values)) +
  geom_histogram(binwidth=5, fill="pink", color="black")
```



## Plots: `ggplot2` density plot

```{r}
ggplot(data=df_IQ, aes(x=IQ_values)) +
  geom_histogram(aes(y=..density..), binwidth=5, fill="pink", 
                 color="black", alpha=.50) +
  geom_density(fill="firebrick", alpha = .50) +
  labs(title="IQ histogram and density", x = "IQ")
```


## Plots: `ggplot2` density plot with mean

```{r}
ggplot(data=df_IQ, aes(x=IQ_values)) +
  geom_density(fill="firebrick", alpha = .30) +
  geom_vline(aes(xintercept = 130), size=3, col="blue") +
  geom_vline(aes(xintercept = 70), size=3, col="firebrick") +
  geom_vline(aes(xintercept = 100), size=3, col="yellow") + 
  geom_vline(aes(xintercept = mean(IQ_values)), size=1, col="orange") +
  labs(title="IQ density with annotations", x="IQ")
```


## Plots: `ggplot2` density plot with normal distribution

```{r}
ggplot(data=df_IQ, aes(x=IQ_values)) +
  geom_density(fill="yellow", alpha = .30) +  
  stat_function(fun = dnorm, n = 101, args = list(mean = 100, sd = 15), 
                col="firebrick", size=2) +
  stat_function(fun = dnorm, n = 101, args = list(mean = mean(IQ_values), sd = sd(IQ_values)), 
                col="orange", size=2) + 
  xlim(50, 150)
```

## Plots: `ggplot2` histogram with normal distribution

```{r}
ggplot(data=df_IQ, aes(x=IQ_values)) +
  geom_histogram(aes(y=..density..), binwidth=10, fill="pink", 
                 color="black", alpha=.50) +  
  stat_function(fun = dnorm, n = 101, args = list(mean = 100, sd = 15), 
                col="blue", size=1)
```

## Plots: Plotting different groups
We will add a group variable `school` and simulate data:

```{r}
IQ_1 = rnorm(n=100, mean=105, sd=10) %>% round()
IQ_2 = rnorm(n=100, mean=120, sd=16) %>% round()
IQ_3 = rnorm(n=100, mean=95, sd=8) %>% round()
df_1 = data.frame(school="school1", IQ=IQ_1)
df_2 = data.frame(school="school2", IQ=IQ_2)
df_3 = data.frame(school="school3", IQ=IQ_3)
df_schools = rbind(df_1, df_2, df_3)
```

## Plots: `ggplot2` density plot with 2 groups
```{r}
ggplot(data=df_schools, aes(x=IQ, fill=school)) +
  geom_density(alpha=.50)
```

## Plots: `ggplot2` barplot

```{r}
ggplot(data=df_schools, aes(x=school, y=IQ, fill=school)) +
  geom_bar(stat="summary", fun="mean") +
  ylim(0, 130)
```

```{r echo=FALSE, include=FALSE}
# cleanup here
```
