---
title: "Introduction to Statistics with R"
subtitle: "Session R04: t-Test"
author: "Marvin Schmitt"
output:
  beamer_presentation:
    slide_level: 2
    toc: no
    theme: Berlin
    includes:
      in_header: ../../config-files/notes-config.tex
header-includes: \setbeamertemplate{footline}[]{}
---




```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(fig.height = 3, fig.width = 6)

library(tidyverse)
library(ggplot2)
library(ggpubr)
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
set.seed(42)
options(warn=-1)
```

## The example data set

This example data set contains depression measures before (`MZP1`) and after (`MZP2`) a novel therapy.

```{r}
df = read.csv("R04_notes_dataset.csv")
nrow(df)
colnames(df)
```

## Data formats: `wide` and `long`

Data sets can come in `wide` format:

```{r echo=FALSE}
df %>% head(3)
```

or `long` format:

```{r echo=FALSE}
df %>% 
  gather(., mzp, depr, MZP1:MZP2) %>% 
  filter(ID %in% 1:3) %>% 
  arrange(ID)
```

## Data formats: conversion `wide` to `long`

```{r}
df %>% head(3)

df %>% 
  gather(., mzp, depr, MZP1:MZP2) -> df_long

df_long %>% arrange(ID) %>% head(6)
```

## Data formats: conversion `long` to `wide`

```{r}
df_long %>% arrange(ID) %>% head(6)

df_long %>%
  spread(., mzp, depr) -> df_wide

df_wide %>% head(3)
```


## `ggplot2`: Barplot with error bars

```{r}
ggplot(data = df_long, aes(y=depr, x=mzp, fill=mzp)) +     # long format!
  stat_summary(fun = mean, geom = "bar", width=0.5) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width=0.2)
```

## t-test syntax

The function `t.test()` implements several versions of the $t-$test:

- One sample:
  - The first argument is the data, `mu` is the constant (aka. $\lambda$)
  - Example: `t.test(IQ_values, mu=100)`
- Two samples:
  - The first two arguments are $x_1$ and $x_2$
  - `paired` controls if the samples are paired (`TRUE`) or independent (`FALSE`, default)
  - `var.equal` controls if variances are assumed to be equal
  - Example: `t.test(x1, x2, var.equal=TRUE)`
  
## Effect size calculations

The library `effsize` provides the function `cohen.d()` to estimate the effect size of a $t-$test:

```{r echo=FALSE}
library(effsize)
```

```{r, eval=FALSE}
library(effsize)
cohen.d(IQ_values, NA, mu=100)  # no second sample -> NA
cohen.d(x1, x2)
```

If we want a shorter output with only the estimated effect size $d$, we can extract that information with the `$` operator:

```{r eval=FALSE}
effect_size_result = cohen.d(x1, x2)
effect_size_result$estimate   # print only the estimate of d

cohen.d(x1, x2)$estimate   # short form
```

## t-test (one sample)

We test whether the average depression score `depr` at $T_1$ differs from $\lambda=19$:

$H_0: \mu = 19, \quad H_1:\mu\neq 19$

```{r}
t.test(df_wide$MZP1, mu=19)

cohen.d(df$MZP1, NA, mu=19)$estimate
```


## t-test (paired variables)

We test whether the patient's depression scores `depr` _change_ between $T_1$ and $T_2$:


$H_0: \bar{x_d} = 0, \quad H_1:\bar{x_d}\neq 0$

```{r}
t.test(df_wide$MZP1, df_wide$MZP2, paired=TRUE)
```

***

We test whether the patient's depression scores `depr` _decrease_ between $T_1$ and $T_2$:


$H_0: \bar{x_d}<0, \quad H_1:\bar{x_d}>0$

```{r}
t.test(df_wide$MZP2, df_wide$MZP1, paired=TRUE, alternative="less")
```

## t-test (two sample)

We test whether the means of $x_1$ and $x_2$ differ:

$H_0: \bar{x_1}=\bar{x_2}, \quad H_1:\bar{x_1}\neq \bar{x_2}$

```{r}
x1 = rnorm(n=20, mean=98, sd=15)
x2 = rnorm(n=20, mean=100, sd=15)
t.test(x1, x2, var.equal = TRUE)

cohen.d(x1, x2)$estimate
```


```{r echo=FALSE, include=FALSE}
# cleanup here
```
