---
title: "Introduction to Statistics with R"
subtitle: "Session R05: ANOVA"
author: "Marvin Schmitt"
output:
  beamer_presentation:
    slide_level: 2
    toc: no
    theme: Berlin
    includes:
      in_header: ../../config-files/notes-config.tex
header-includes: \setbeamertemplate{footline}[]{}
---




```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(fig.height = 3, fig.width = 6)

library(tidyverse)
library(ggplot2)
library(ggpubr)
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
set.seed(42)
options(warn=-1)
```

## The example data set

This example data set contains scores on a sustainability scale for three groups, each with a distinct diet: meat-eating (`meat`), vegetarian (`vegetarian`) and plant-based (`vegan`). Furthermore, each participant's self-reported gender is denoted. 

**Tip:** Use the parameter `stringsAsFactors` to interpret character strings as factors automatically:

```{r}
df = read.csv("R05_notes_dataset.csv", stringsAsFactors = TRUE)
str(df)

```

***

We can change the **ordering** of the **factor levels**:

```{r}
df$diet = factor(df$diet, levels=c("vegan", "vegetarian", "meat"))
```

## Missing data

In the `R` language, missing data are denoted as `NA` (not available). That's a dedicated symbol. It's not to be confused with `NaN` (not a number) that describes impossible data (e.g. from divison by 0).

```{r}
head(df)
```

***

Some functions do not work properly when `NA`s are present:

```{r}
mean(df$sustainability)
```

In many cases, these functions implement the argument `na.rm=TRUE` (= remove `NA` values):

```{r}
mean(df$sustainability, na.rm=TRUE)
```

*** 

There are different strategies for missing values:

- Delete the data
  - Delete the entire observation (row)
  - Exclude the observation only from those analyses where the missing value would be required
- Substitute the missing value with a _typical_ value, like ...
  - the mean $\bar{x}$
  - the median
  - the mode
  - ...
- Copy the last observation
- Estimate the missing value
  - Regression
  - Multiple imputation
  - ...
  
Introductory reading: <https://doi.org/10.4097/kjae.2013.64.5.402>

***

We can delete all rows that contain **any** missing data cell with `na.omit()`. That's simple but wasteful in applications where data are valuable.

```{r}
df = na.omit(df) 
```


## Visualization: one-way ANOVA

For a one-way ANOVA (one grouping variable), we can simply use a bar plot with error bars:

```{r out.width="70%"}
df %>% filter(gender=="male") %>% 
  ggplot(., aes(y=sustainability, x = diet)) +
  stat_summary(fun = mean, geom = "bar", width=0.5, fill="#4E84C4") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width=0.2)
```


## Visualization: Two-way ANOVA

For two factors, we can use a bar plot with different `fill`s. The bars and error bar need to be _dodged_ a little:
```{r out.width="70%"}
ggplot(df, aes(y=sustainability, x = diet, fill=gender)) +
stat_summary(fun=mean, geom="bar", width=0.5, position=position_dodge(0.55)) +
stat_summary(fun.data=mean_se, geom="errorbar", width=.2, 
             position=position_dodge(.55))
```

***

```{r}
ggplot(df, aes(y=sustainability, x = gender, group=diet, color=diet)) +
  stat_summary(fun = mean, geom = "pointrange", position=position_dodge(0.1))+
  stat_summary(fun=mean, geom="line", position=position_dodge(0.1)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width=0.1, 
               position=position_dodge(0.1))
```

## Numerical ANOVA computation

The `afex` package implements convenient functions to compute ANOVAs. We will use the function `aov_ez()`:

```{r, message = FALSE}
library(afex)
aov_ez(id = "ID",             # identifier for subjects
       dv = "sustainability", # dependent variable (y)
       between = "gender",    # between factor (group)
       data = df)             # data frame
```

***

**One-way ANOVA**: Effect of `gender`

```{r, warnings=FALSE}
m = aov_ez(id = "ID", dv = "sustainability", 
           between = "gender", data = df) 
summary(m)
```

***

**One-way ANOVA**: Effect of `diet`

```{r, message = FALSE}
m = aov_ez(id = "ID", dv = "sustainability", 
           between = "diet", data = df) 
summary(m)
```

***

**Two-way ANOVA**: Effect of `gender` and `diet`

```{r, message = FALSE}
aov_ez(id = "ID", dv = "sustainability", 
       between = c("gender", "diet"), data = df) %>% summary()
```

```{r echo=FALSE, out.width="70%"}
ggplot(df, aes(y=sustainability, x = diet, group=gender, color=gender)) +
  stat_summary(fun = mean, geom = "pointrange", position=position_dodge(0.1))+
  stat_summary(fun=mean, geom="line", position=position_dodge(0.1)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width=0.1, 
               position=position_dodge(0.1))
```

***

```{r}
df_nicotine = read.csv("R05_notes_dataset_nicotine.csv", stringsAsFactors=TRUE)
```
```{r echo=FALSE}
ggplot(df_nicotine, aes(y=nicotine_conc, x = time, group=nationality, color=nationality)) +
  stat_summary(fun = mean, geom = "pointrange", position=position_dodge(0.1))+
  stat_summary(fun=mean, geom="line", position=position_dodge(0.1)) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width=0.1, 
               position=position_dodge(0.1))
```

***

**Within-subject ANOVA** for factor `time`:

```{r, message = FALSE}
aov_ez(id = "ID", dv = "nicotine_conc", 
       within = "time", data = df_nicotine) %>% summary()
```

***

```{r, message = FALSE}
aov_ez(id = "ID", dv = "nicotine_conc", between="nationality",
       within = "time", data = df_nicotine) %>% summary()
```


```{r echo=FALSE, include=FALSE}
# cleanup here
```
