---
title: "Introduction to Statistics with R"
subtitle: "Session R07: Conclusion"
author: "Marvin Schmitt"
output:
  beamer_presentation:
    slide_level: 2
    toc: no
    theme: Berlin
    includes:
      in_header: ../../config-files/notes-config.tex
header-includes: \setbeamertemplate{footline}[]{}
---




```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(mlbench)
```

## Admin: Certificates

- What kind of certificate do you need?
- Who needs ECTS credits?


## Recap: syllabus

Date | Topic | Location
--- | --- | ---
19.10.2021, 09:15 - 12:45 | Kickoff and setup | in-person   
26.10.2021, 09:15 - 10:45 | Basics and diagrams | Online
02.11.2021, 09:15 - 10:45 | Correlation | Online
09.11.2021, 09:15 - 10:45 | Regression | Online
16.11.2021, 09:15 - 10:45 | t-Test | Online
23.11.2021, 09:15 - 10:45 | ANOVA | Online
30.11.2021, 09:15 - 11:15 | Eye-Tracking | Online
30.11.2021, 11:30 - 12:30 | Conclusion | Online


## Potential applications: Think, pair, share!

**How can and will I apply the content of this seminar to my own research?**

1. **Think** about the question alone for *3 minutes*.
2. **Pair** up with someone and talk about the question for *15 minutes*.
3. **Share** one highlight of your pair-talks with the entire group.


## Feedback

Looking back at the seminar retrospectively:

- What do you pack in your **suitcase**?
- What do you toss in the **bin**?
- Anything that you want to let me know?

***

```{r}
text = c("!", "goodbye", "you", "Thank" , "and")
indices = c(4, 3, 5, 2, 1)
for (idx in indices){
  print(text[idx])
}
```

```{r, echo=FALSE, out.width="60%", fig.align = 'center'}
d = mlbench.smiley(n=1000, sd1 = 0.1, sd2 = 0.1)
data.frame(
  x = d$x[, 1],
  y = d$x[, 2],
  class = d$classes
) %>% 
  filter(class != 3) %>% 
  ggplot(., aes(x = x, y=y, color=class)) +
  geom_point() + theme_bw() + theme(legend.position="none")
```

