---
title: "Introduction to Statistics with R"
subtitle: "Session R02: Correlation"
author: "Marvin Schmitt"
output:
  beamer_presentation:
    slide_level: 2
    toc: no
    theme: Berlin
    includes:
      in_header: ../../config-files/notes-config.tex
header-includes: \setbeamertemplate{footline}[]{}
---




```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(fig.height = 3, fig.width = 6)

library(tidyverse)
library(ggplot2)
library(ggpubr)
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
set.seed(42)
options(scipen=10000)
options(warn=-1)
```

## The example data set

We analyze structures in the **Chicago Face Database** (Ma et al., 2015). Each row refers to a portrait which was rated with respect to different categories by a sample of raters.

```{r}
df = read.csv("R02_notes_dataset.csv")
nrow(df)
colnames(df)
```

## Scatterplots

```{r warnings=FALSE}
ggplot(df, aes(x=Masculine, y=Attractive))+
geom_point() +
ylim(1,5) + xlim(1,5)
```

***

**Digression I: `filter`**

```{r}
df %>% filter(Gender=="M")
```

***

**Digression II: `select`**
```{r}
df %>% 
  select(Attractive)
```

***

```{r}
df %>% filter(Gender=="M") %>% 
  ggplot(., aes(x=Masculine, y=Attractive))+
  geom_point() +
  ylim(1,5) + xlim(1,5)
```

***
```{r}
df %>% filter(Gender=="F") %>% 
  ggplot(., aes(x=Masculine, y=Attractive))+
  geom_point() +
  ylim(1,5) + xlim(1,5)
```

## Scatterplots: Group variables

```{r}
ggplot(df, aes(x = Masculine, y = Attractive, color=Gender))+
  geom_point() +
  ylim(1, 5) + xlim(1, 5)
```

## Scatterplots: Print correlation coefficients

```{r}
ggplot(df, aes(x=Masculine, y=Attractive))+
geom_point() +
ylim(1, 6) + xlim(1, 5) +
stat_cor(p.accuracy = 0.01)
```



***

```{r}
ggplot(df, aes(x=Masculine, y=Attractive, color=Gender))+
geom_point() +
ylim(1, 6.5) + xlim(1, 5) +
stat_cor(p.accuracy = 0.01)
```

## Correlation matrices: Computation


```{r}
df %>% 
  select(Angry, Disgusted, Happy, Sad, Surprised, Attractive, Threatening) %>% 
  cor() %>% 
  round(3)
```

## Correlation matrices: Visualization

We can visually encode the entries in a **correlation matrix** while maintaining the matrix structure:

- Color coding
- Circles
- Ellipses
- Correlation coefficients $r_{xy}$
- ...

We can use the function `corrplot::corrplot` to visualize correlation matrices. 

***

```{r out.width = "50%"}
df %>% 
  select(Angry, Disgusted, Happy, Sad, Surprised, Attractive, Threatening) %>% 
  cor() %>% 
  corrplot::corrplot()
```

***

```{r out.width = "50%"}
df %>% 
  select(Angry, Disgusted, Happy, Sad, Surprised, Attractive, Threatening) %>% 
  cor() %>% 
  corrplot::corrplot(method="ellipse")
```

***

```{r out.width = "50%"}
df %>% 
  select(Angry, Disgusted, Happy, Sad, Surprised, Attractive, Threatening) %>% 
  cor() %>% 
  corrplot::corrplot(method="number", number.cex = 0.7)
```

***

**Important arguments for `corrplot::corrplot`**:

- `method`: cell content
- `tl.col='black'`: black labels
- `cl.pos='n'`: remove colorbar
- `type`: display entire/lower/upper matrix
- `number.cex`: size of numbers if `method=number`


## Testing assumptions 

In order to ease interpretation of a correlation coefficient $r_{xy}$, the variables $X, Y$ should be:

- **metric**,
- **normal**, and
- **homoscedastic**.

*** 

We can use the function `shapiro.test(x)` to test `x` for **normality**.

**Interpretation:**

- if $p\geq.05$: normality assumption not violated
- if $p<.05$: normality assumption violated

```{r}
shapiro.test(df$Trustworthy)
shapiro.test(df$Attractive)
```

***

The scatterplot allows for a visual inspection of **homoscedasticity**:

```{r echo=FALSE}
x = seq(1, 100, length.out=500)
y = rnorm(n=500, mean=0.5*x, sd=x/10)

data.frame(x,y) %>% 
  ggplot(aes(x=x, y=y)) +
  geom_point()
```

***

**Another caveat:**

```{r echo=FALSE}
x = seq(1, 10, length.out=100)
y = rnorm(n=100, mean=(x-5.5)**2, sd=0.1)
z = rnorm(n=100, mean=10-(x-5.5)**2, sd=1)
```

```{r echo=FALSE}
data.frame(x,y,z) %>% 
  ggplot(aes(x=x)) +
  geom_point(aes(y=y), col="red") + 
  stat_cor(aes(y=y), col="red", p.accuracy = 0.01, label.x = 3) +
  ylim(-10, 20)
```

***

```{r echo=FALSE}
data.frame(x,y,z) %>% 
  ggplot(aes(x=x)) +
  geom_point(aes(y=y), col="red") + 
  geom_point(aes(y=z), col="blue") +
  stat_cor(aes(y=y), col="red", p.accuracy = 0.01, label.x = 3) +
  stat_cor(aes(y=z), col="blue", p.accuracy = 0.01, label.x = 6)
```


*** 

```{r echo=FALSE}
x = seq(1, 10, length.out=50)
y = rnorm(n=50, mean=(x-5.5)**2, sd=0.5)
w = rnorm(n=50, mean=(x-5.3)**2, sd=0.5)
z = rnorm(n=50, mean=(x-5.0)**2, sd=0.5)

data.frame(x,y,w,z) %>% 
  ggplot(aes(x=x)) +
  geom_point(aes(y=y), col="darkorange") + 
  stat_cor(aes(y=y), col="darkorange", p.accuracy = 0.01, label.x = 3, label.y=24)+
  geom_point(aes(y=w), col="forestgreen") + 
  stat_cor(aes(y=w), col="forestgreen", p.accuracy = 0.01, label.x = 3, label.y=20)+
  geom_point(aes(y=z), col="brown") + 
  stat_cor(aes(y=z), col="brown", p.accuracy = 0.01, label.x = 3, label.y=16)

```


***

```{r echo=FALSE}
x = seq(1, 10, length.out=100)
y = rnorm(n=100, mean=sin(x), sd=0.1)
z = rnorm(n=100, mean=cos(x), sd=0.1)


data.frame(x,y,z) %>% 
  ggplot(aes(x=x)) +
  geom_point(aes(y=y), col="red") + 
  geom_point(aes(y=z), col="blue") +
  ylim(-2, 2) +
  stat_cor(aes(y=y), col="red", p.accuracy = 0.01, label.x = 5) +
  stat_cor(aes(y=z), col="blue", p.accuracy = 0.01, label.x = 0)
```

```{r echo=FALSE, include=FALSE}
# cleanup here
```
