---
title: "Introduction to Statistics with R -- FAQ"
date: "Page updated: `r format.Date(file.mtime('FAQ.Rmd'),'%d.%m.%Y')`"
---

### I cannot load files, such as data sets in `.csv` format
The problem might be that your working directory is not correctly set. Execute the command

```{r eval=FALSE}
getwd()
```

and see if the printed working directory matches your expectation. That is the root of all your file references. 

**Tip:** You can set the working directory to the folder of the current `.R` or `.Rmd` file with:

```{r eval=FALSE}
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
```

### Problems with generating RMarkdown Notebooks
You might have not installed `knitr` which generates the pretty files from your `.Rmd` notebooks. Try to execute the following code to install `knitr`:

```{r eval=FALSE}
install.packages("knitr")
```

### RMarkdown Notebooks won't generate even though `knitr` is installed
The problem might be that there is no sufficient LaTeX distribution installed. LaTeX is used for typesetting. Install a minimum LaTeX distribution with the following code:

```{r eval=FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

